---
title: |
 | ECON 7201
 | Applied Econometrics
subtitle: "Assignment 1"
author: Mamadou Yacine BAH
format:
  pdf: 
   include-in-header: 
    text: |
     \usepackage{fancyhdr}
     \fancypagestyle{style2}{
     \fancyhf{}
     \fancyhead[R]{Assignment 1}
     \fancyhead[L]{ECON 7201}
     \fancyfoot[C]{\thepage}
     \renewcommand{\headrulewidth}{1pt}
     }
      \pagestyle{style2}
execute:
  eval: true
  echo: true
---


```{r}
#| label: setup
#| include: false

#install.packages("tikzdevice")
# Load the tikdevice library
library(tikzDevice)

library(ggplot2)
library(dplyr)

# Set the default graphics size for pdf for the entire document
knitr::opts_chunk$set(fig.width = 6, fig.height = 4)

set.seed(42)
```





\vspace{-1in}

## Due Date

**September 18, 2025** at the start of class

## Directions

Answer all questions. Submit both a PDF and Quarto file to the nexus assignment portal.

# 1. Git and GitHub

(a) Create a GitHub repository called **econ_3201** and connect it to RStudio.
(b) Create a new R project in this newly created directory called **assignment_1**. (Note, you do not have to click "Create git repository" as the directory is contained in a git enabled directory, i.e., **econ_3201**).
(c) Download the assignment PDF and Quarto file the **assignment_1** folder.
(d) Commit and push the changes to your **econ_3201** repository on [GitHub.com](GitHub.com).

# 2. LaTeX

LaTeX is useful for writing math equations and presents them in a neat and orderly way. To write in math mode, wrap your text in `$` for inline text use two `$`s for display (i.e., centered on the page). Some very useful functions include:

-   Fractions:`\frac{}{}`, e.g. `$\frac{1}{2}$` gives $\frac{1}{2}$ and `$$\frac{1}{2}$$` gives: $$\frac{1}{2}.$$
-   Subscripts: `_` gives a subscript, e.g. `$x_1$` gives $x_1$. To include more than one term in the subscript, the items in the subscript must be enclosed by `{}`. E.g. `$x_{1,1}$` gives $x_{1,1}$. (Note that `$x_1,1$` gives $x_1,1$)
-   Exponents: `^`, e.g. `$x^2$` gives $x^2$. `^` can also be used for superscripts in other math functions, including summations and integrals.
-   Aligned: `aligned` neatly aligns multiple lines of an equation. Align is useful when writing multiple steps to solving an equation. To use it in Quarto, write `$$\begin{aligned}...\end{aligned}`. The `&` is used to mark the point where the lines should be aligned. Use `\\` at the end of each line E.g. Â  `$$\begin{aligned}`\
    `x & = 3 + 5\\`\
    `& = 8`\
    `\end{aligned}$$`Â 

Â  Â  Â  gives $$\begin{aligned}
x & = 3 + 5\\
& = 8
\end{aligned}$$\`

-   Summation: `\sum` gives the summation sign, i.e. $\sum$. To include subscripts, use `_` and to use superscripts use `^`, e.g. `$\sum_{i=1}^n$` gives $\sum_{i=1}^n$, which reads as the sum of $i$ equals 1 to $n$.
-   Integral: `\int` gives an integral, i.e. $\int$. To place a lower limit use `_` and to place an upper limit, use `^`, e.g. `$\int_{a}^{b}$` gives $\int_{a}^{b}$.
-   Greek letters: `$\alpha, \beta, \gamma, \Gamma, \delta, \Delta, \epsilon, \varepsilon, \zeta, \eta,\sigma, \Sigma, \theta, \vartheta, \Theta, \iota, \kappa, \lambda, \Lambda, \mu$` gives $\alpha, \beta, \gamma, \Gamma, \delta, \Delta, \epsilon, \varepsilon, \zeta, \eta, \sigma, \Sigma, \theta, \vartheta, \Theta, \iota, \kappa, \lambda, \Lambda, \mu$. (See https://www.overleaf.com/learn/latex/List_of_Greek_letters_and_math_symbols )
-   Accents: `\hat{}`, `\tilde{}`, and `\bar{}`are examples of accents in math mode. E.g. `\hat{Y}`, `\tilde{Y}`, and `\bar{Y}`gives $\hat{Y}$, $\tilde{Y}$, and $\bar{Y}$, respectively.
-   Text: To include text in your equation, i.e. non italicized text, use `\text{}`, e.g. `$x=2\text{ if } y=1$` gives $x=2\text{ if } y=1$.
-   Inequalities: Some mathematical expressions may be written as inequalities, rather than equations. For 'less than' and 'greater than', you can just use the symbol on your keyboard, i.e. `<` and `>`, respectively. For $\leq$, use `$\leq$` and for $\geq$, use `$\geq$`. An important note is that after writing a command, put a space after the command before writing the next term, otherwise you may get an error. E.g. To write $a\leq b$, write `$a\leq b$`, not `$a\leqb$`.

Â 

Re-write the following equations in LaTeX. \`

(a) $E(Y) = y_1p_1 + ... + y_kp_k = \sum_{i=1}^ky_ip_i$

(b) $\sigma_y = Var(Y) = E[(Y - \mu_y)^2] = \sum_{i=1}^k (y_i - \mu_y)^2p_i$

(c) $\hat{\beta} = \frac{\sum_{i=1}^n(y - y_i)(x - x_i)}{\sum_{i=1}^n(x -x_i)^2}$

(d) $P(a \leq Y \leq b) = \int_{a}^{b} f_Y(y)dy$

(e) $\hat{g}(x) = \frac{\frac{1}{nh} \sum_{i=1}^n y_ik (\frac{x_i - x}{h})}{\frac{1}{nh} \sum_{i=1}^n k(\frac{x_i - x}{h})}$

# 3. R

## 3.1. Assignment

Note: When creating variables based on equation, separate each element in the equation with the appropriate arithmetic symbol. E.g., to compute $x(y-2)$ in R, you would have to type `x*(y-2)`. `x(y-2)`, with not arithmetic symbol between $x$ and the left bracket would result in an error.

(a) In statistics, $n$ is often used to denoted the sample size. Set the number of observations $n=1000$.

```{r}
n <- 1000
```

(b) Generate two random variables, $u_1\sim U(0,1)$ and $u_2\sim U(0,1)$ with $n/2=500$ observations. That is, create two variables that follow a uniform distribution between 0 and 1 that each have 500 observations. In R, we can create random uniform variables using the `runif(k,min,max)` function, where `k` is number of observations, `min` is the minimum value, and `max` is the maximum value. The default values for `min` and `max` are 0 and 1, respectively. Type `?runif` into your console to learn more.

```{r}
u_1 <- runif(n/2, 0, 1)

u_2 <- runif(n/2, 0, 1)
```

(c) Generate two variables $z_1$ and $z_2$ that take on the following values: $$ z_1 = \sqrt{âˆ’2\ln(ð‘¢1)} Ã—cos(2ðœ‹ð‘¢2)$$ and $$z_2=\sqrt{-2\ln(u_1)}\times\text{sin}(2\pi u_2).$$ In R, $\sqrt{\text{  }}$ is computed using `sqrt()`, $\ln$ is computed using `log()`, $\text{cos}$ is computed using `cos()`, and $\text{sin}$ is computed using `sin()`.

```{r}

z_1 <- sqrt(-2*log(u_1))*cos(2*pi*u_2)

z_2 <- sqrt(-2*log(u_1))*cos(2*pi*u_2)

```

(d) Generate a vector $z=[z_1,z_2]$

```{r}

z <- c(z_1, z_2)

```

(e) Generate two variables $\mu$ (spelled mu) and $\sigma$ (spelled sigma). Set $\mu=5$ and $\sigma=2$.

```{r}

mu <- 5

sigma <- 2
```

(f) Generate a variable $x=\mu+\sigma\times z$

```{r}

x <- mu + sigma * z
```

(g) Calculate the mean of $x$, using `mean()` and the standard deviation of $x$ using `sd()`.

```{r}
mean(x)
sd(x)
```

(h) Use the following code to plot a histogram of $x$ with the normal distribution curve. \`

```{r}
#| fig-cap: "Histogram of x"
#| dependson: q1
#| eval: true
#| echo: true
#| dev: pdf


hist(x,
     freq = FALSE,
     ylab = "Density",
     xlab = "x",
     main = "Histogram")
     
     
curve(dnorm(x, mean = mu, sd = sigma),
      col = "red", lwd = 2, add = TRUE)
```

## 3.2. Data frames and Indexing

A data frame in R is a table-like data structure used to store data in rows and columns, similar to a spreadsheet or a database table. It is one of the most commonly used structures for storing datasets in R.\
\
@tbl-hlth displays the total health expenditure by use of funds in Canada from 1975 to 2022. The data is stored in the `data.frame` called `df`.

```{r}
#| label: tbl-hlth
#| eval: true
#| echo: true
#| tbl-cap: "Total health expenditure by use of funds, in millions of current dollars, Canada, 1975 to 2022 (Source: CIHI National Health Expenditure Trends)"
cihi <- read.csv("hlthexp.csv")

df <- data.frame(Year = cihi$Year,
                 Hospitals = cihi$Hospitals,
                 Physicians = cihi$Physicians,
                 "Other Services" = cihi$Other.Institutions,
                 Dental = cihi$Other.Professionals..Dental.Services,
                 Vision = cihi$Other.Professionals..Vision.Care.Services,
                 "Other Professionals" = cihi$Other.Professionals..Other.Services,
                 check.names = FALSE)

knitr::kable(df, 
             align = c(rep('c',7)),
             format.args = list(big.mark = ","))
```

(a) Determine if there are any missing values for the variable `Hospitals`.

```{r}

is.na(df$Hospitals)

```

# there is no missing values for the variable "Hospitals"

(b) Add a variable called "Total Other Services" to the data frame `df`, where $$\text{Total Other Services} = \text{Dental}+\text{Vision}+\text{Other Professional}.$$ To add a newly created variable to a data frame use the syntax `dataframe$varname <- expression`.

```{r}

df$`Total Other Services` <- df$Dental + df$Vision + df$`Other Professionals`
```

(c) Are there any years for which Total Other Professionals

(d) Another way to add a variable to a data frame is to simply create a new data frame and append the new variable to it. Note: we can use the same data frame name. I.e., `df<-data.frame(df,newvarname = newvar)`. Add the variable "Prescription Drugs" to the `df` data frame using the append method, where presricption drugs is named "Prescribed.Drugs" in the `cihi` data.frame.

```{r}

df <- data.frame(df, Prescribed.Drugs = cihi$Prescribed.Drugs)
```

sum(Hospitals) (e) Using a single R command, determine the expenditure on hospitals in 1983.

```{r}

df$Hospitals[df$Year == 1983]
```

(a) Using a singe R command, list the expenditures by year for 2012-2022.

```{r}

df[df$Year %in% 2012:2022, ]
```

## 3.3 Other useful R commands.

Load the `mpg` dataset from the `ggplot2` package using `mpg <-ggplot2::mpg`. (Be sure to install the `gglot2` package before you start.)

\`

```{r}
library(ggplot2)
mpg <-ggplot2::mpg
```

(a) Subset the data to include only observations from 2008. Search `?subset` in the console. (a) Calculate the maximum and minimum miles per gallon in city limits (`cty`). Seach `?min` in the console.

```{r}
mpg <- subset(mpg, year >= 2008)

```

```{r}

max(mpg$cty)

min(mpg$cty)
```

(b) Estimate the average miles per gallon within city limits for cars produced in 2008 using the formula $$\text{Average mpg}=\frac{\sum_{i=1}^n \text{cty}_i}{n}.$$ Recall that $n$ is the number of observations. Search `?length` in the console.

```{r}

`Average mpg` = sum(mpg$cty) / length(mpg$cty)

`Average mpg`
```

(c) Estimate the average miles per gallon within city limits for cars produced in 2008 using the `mean()` function.

```{r}

`Averag mpg 2008 city` = mean(mpg$cty)
`Averag mpg 2008 city`
```

(d) Create a variable called `compact`, which takes a value of 1 if the vehichle is a compact and 0 otherwise. Search `?ifelse` in the console.

```{r}

mpg$compact <- ifelse(mpg$class == "compact", 1, 0)
```

(e) Estimate the average miles per gallon within city limits for compact cars. (You may use whichever method you prefer).

```{r}

`Average mpg city compact` = sum(mpg$compact) / length(mpg$compact)

`Average mpg city compact`
```

(f) Create a simple scatter plot with city mpg (`cty`) on the x-axis and highway mpg (`hwy`) on the y-axis. Search `?plot` and choose "Generic X-Y Plotting".

```{r}
plot(x    = mpg$cty,
     y    = mpg$hwy,
     main = "simple scatter plot",
     xlab = "cty",
     ylab = "hwy",
     col  = "blue",
     pch  = 10)
```

        
(i) Change the x-axis label using the option `xlab = "City MPG"` and change the y-axis label using the option `ylab = "Highway MPG"`.

```{r}
#| label: fig-mpgplot
#| fig-cap: "mpg vs hwy"

plot(x    = mpg$cty,
     y    = mpg$hwy,
     xlab = "City MPG",
     ylab = "hwy",
     col  = "blue")
```

        
(ii) Add the caption "City Versus Highway Fuel Efficiency (MPG)"

```{r}
plot(x    = mpg$cty,
     y    = mpg$hwy,
     main = "City Versus Highway Fuel Efficiency (MPG)",
     xlab = "City MPG",
     ylab = "hwy",
     col  = "blue",)
```



(iii) Cross reference the figure and add the text "Figure 1 shows the fuel efficiency for city driving versus highway driving". 

@fig-mpgplot shows the fuel efficiency for city driving versus highway driving


We used the `rnorm()` command